Competing Bandits: \\ The Perils of Exploration under Competition

### 2019-02-14 paper/content/introduction.tex
edits in the abtract
light edits on intro

\begin{figure}
\begin{center}
\begin{tikzpicture}
      \draw[->] (-.5,0) -- (7,0) node[right] {};
      \draw[->] (0,-.5) -- (0,3) node[above] {Better algorithms};
      \draw[scale=0.6,domain=0.7:9.8,smooth,variable=\x,blue, line width=0.3mm] plot ({\x},{4.5 - 0.18 * (\x - 5.25)^2});
     \node[below] at (0.7, -0.22) {\footnotesize monopoly};
     \node[below] at (3, -0.2) {\footnotesize incumbent};
     \node[below] at (4.5, -0.22) {\footnotesize duopoly};
     \node[below] at (6, -0.2) {\footnotesize entrant};
 \end{tikzpicture}
 \caption{A stylized ``inverted-U relationship" between strength of competition and ``level of innovation".}
\label{fig:inverted-U}
\end{center}
\end{figure}

@@

\xhdr{Main findings.}
We find that in the permanent duopoly, competition incentivizes firms to choose the ``greedy algorithm", and even more so if the firm is a late arriver in a market. This algorithm also prevails under monopoly, simply because it tends to be easier to deploy. Whereas the incumbent in the temporary monopoly is incentivized to deploy a more advanced exploration algorithm that performs better in the long run. As a result, we find that consumer welfare is highest under temporary monopoly and that even increasing the number of firms in the market past a duopoly makes consumers weakly worse off.

Interpreting the adoption of better algorithms as ``innovation", our findings can be framed in terms of the ``inverted-U relationship" between competition and innovation (see Figure~\ref{fig:inverted-U}). This relationship -- too little or too much competition is bad for innovation, but intermediate levels of competition tend to be better -- is a familiar theme in the economics literature, dating back to \cite{Schumpeter-42}. However, the ``inverted-U relationship'' is driven by different aspects in our model than the ones in the existing literature in economics. In our case, the barriers for innovations arise entirely from the reputational consequences of exploration in competition, as opposed to the R\&D costs (which is the more standard cause in prior work).

\xhdr{Additional findings.}
We investigate the ``first-mover advantage" phenomenon in more detail. Being first in the market gives free data to learn from (a ``data advantage") as well as a more definite, and possibly better reputation compared to an entrant (a ``reputation advantage"). We run additional experiments so as to isolate these two effects. We find that the data advantage alone can lead to a significant advantage under competition. In fact, even a small ``head start'' can get amplified into a large incumbency advantage under competition. This effect is most pronounced when the incumbent commits to a more advanced bandit algorithm. These findings suggest that competition dynamics are pertinent to the anti-trust considerations of data as a barrier to entry.

\OMIT{ %%%%%%
\swedit{Beyond the welfare implications, we also investigate the barrier to entry that follows from the incumbent's data advantage.  We run additional experiments that attempt to isolate the impact of the data advantage and reputation advantage. We find that the data advantage from the incumbent's ``small head start'' can lead to a significant incumbency advantage, especially when the incumbent commits to a more advanced bandit algorithm. Our results show that even a small degree of data asymmetry can be amplified to be a larger market advantage through competition dynamics. \gadelete{This is in contrast to the data advantage in a static environment such as discussed in \cite{varian2018artificial,bajari2018impact}, in which a small amount of additional data can only provide limited improvement in the predictive accuracy.} The effect is stronger when the incumbent commits to a more advanced bandit algorithm due to
}} %%%%%%
% in a competitive environment the relative differences in the data
% possessed by the firms may lead to a large competitive benefit due
% to the feedback loops discussed previously. In our model this effect
% is amplified when the incumbent commits to a more advanced bandit
% algorithm since, in the incumbency period, the firm gathers data on
% possible actions more efficiently leading to the relative gap upon
% entry to be larger. This is important when exploration can have
% larger reputational consequences.

>>>>

@@

\xhdr{Main findings.}
We find that in the permanent duopoly, competition incentivizes firms to choose the ``greedy algorithm", and even more so if the firm is a late arriver in a market. This algorithm also prevails under monopoly, simply because it tends to be easier to deploy. Whereas the incumbent in the temporary monopoly is incentivized to deploy a more advanced exploration algorithm that performs better in the long run. As a result, we find that consumer welfare is highest under temporary monopoly and that even increasing the number of firms in the market past a duopoly makes consumers weakly worse off.\swcomment{why ``even''?}

Interpreting the adoption of better algorithms as ``innovation", our
findings can be framed in terms of the ``inverted-U relationship"
between competition and innovation (see Figure~\ref{fig:inverted-U}),
\swedit{a well-established concept in the economics literature, dating
  back to \cite{Schumpeter-42}.} \swedit{In this relationship, too
  little or too much competition is bad for innovation, but
  intermediate levels of competition tend to be better.} However, the
``inverted-U relationship'' is driven by different aspects in our
model than the ones in the existing literature in economics. In our
case, the barriers for innovations arise entirely from the
reputational consequences of exploration in competition, as opposed to
the R\&D costs (which is the more standard cause in prior work).

\begin{figure}[t]
\begin{center}
\begin{tikzpicture}
      \draw[->] (-.5,0) -- (7,0) node[right] {};
      \draw[->] (0,-.5) -- (0,3) node[above] {Better algorithms};
      \draw[scale=0.6,domain=0.7:9.8,smooth,variable=\x,blue, line width=0.3mm] plot ({\x},{4.5 - 0.18 * (\x - 5.25)^2});
     \node[below] at (0.7, -0.22) {\footnotesize monopoly};
     \node[below] at (3, -0.2) {\footnotesize incumbent};
     \node[below] at (4.5, -0.22) {\footnotesize duopoly};
     \node[below] at (6, -0.2) {\footnotesize entrant};
 \end{tikzpicture}
 \caption{A stylized ``inverted-U relationship" between strength of competition and ``level of innovation".}
\label{fig:inverted-U}
\end{center}
\end{figure}

\xhdr{Additional findings.}
We investigate the ``first-mover advantage" phenomenon in more detail. Being first in the market gives free data to learn from (a ``data advantage") as well as a more definite, and possibly better reputation compared to an entrant (a ``reputation advantage"). We run additional experiments so as to isolate these two effects. We find that the data advantage alone can lead to a significant advantage under competition. In fact, even a small ``head start'' can get amplified into a large incumbency advantage under competition. This effect is most pronounced when the incumbent commits to a more advanced bandit algorithm. These findings suggest that competition dynamics are pertinent to the anti-trust considerations of data as a barrier to entry.
\swcomment{should we say ``reputation advantage'' alone does not give much advantage?}
\OMIT{ %%%%%%
\swedit{Beyond the welfare implications, we also investigate the barrier to entry that follows from the incumbent's data advantage.  We run additional experiments that attempt to isolate the impact of the data advantage and reputation advantage. We find that the data advantage from the incumbent's ``small head start'' can lead to a significant incumbency advantage, especially when the incumbent commits to a more advanced bandit algorithm. Our results show that even a small degree of data asymmetry can be amplified to be a larger market advantage through competition dynamics. \gadelete{This is in contrast to the data advantage in a static environment such as discussed in \cite{varian2018artificial,bajari2018impact}, in which a small amount of additional data can only provide limited improvement in the predictive accuracy.} The effect is stronger when the incumbent commits to a more advanced bandit algorithm due to
}} %%%%%%
% in a competitive environment the relative differences in the data
% possessed by the firms may lead to a large competitive benefit due
% to the feedback loops discussed previously. In our model this effect
% is amplified when the incumbent commits to a more advanced bandit
% algorithm since, in the incumbency period, the firm gathers data on
% possible actions more efficiently leading to the relative gap upon
% entry to be larger. This is important when exploration can have
% larger reputational consequences.
